\documentclass{article}

\usepackage{fancyhdr} % Required for custom headers
\usepackage{lastpage} % Required to determine the last page for the footer
\usepackage{extramarks} % Required for headers and footers
\usepackage[usenames,dvipsnames]{color} % Required for custom colors
\usepackage{graphicx} % Required to insert images
\usepackage{listings} % Required for insertion of code
\usepackage{courier} % Required for the courier font
\renewcommand{\_}{\char`_}

% Margins
\topmargin=-0.45in
\evensidemargin=0in
\oddsidemargin=0in
\textwidth=6.5in
\textheight=9.0in
\headsep=0.25in

\linespread{1.1} % Line spacing

\lstset{language=C,
                basicstyle=\ttfamily,
                keywordstyle=\color{blue}\ttfamily,
                stringstyle=\color{red}\ttfamily,
                commentstyle=\color{Plum}\ttfamily,
                morecomment=[l][\color{magenta}]{\#}
}


% Set up the header and footer
\pagestyle{fancy}
\lhead{Group 18} % Top left header
\chead{Task 0: Alarm Clock} % Top center head
\rhead{\firstxmark} % Top right header
\lfoot{\lastxmark} % Bottom left footer
\rfoot{Page\ \thepage\ of\ \protect\pageref{LastPage}} % Bottom right footer
\renewcommand\headrulewidth{0.4pt} % Size of the header rule
\renewcommand\footrulewidth{0.4pt} % Size of the footer rule

\setlength\parindent{0pt} % Removes all indentation from paragraphs

%----------------------------------------------------------------------------------------
%	TITLE PAGE
%----------------------------------------------------------------------------------------

\title{
\vspace{2in}
\textmd{\textbf{Task 1: Scheduling}}\\
\normalsize\vspace{0.1in}\small{Due\ on\ Tuesday,\ February\ 9,\ 2016}\\
\vspace{0.1in}\large{\textbf{Pintos Group 18}}
\vspace{3in}
}

\author{Corentin Herbinet, Ignacio Navarro, Vinothan Shankar, William Springsteen}
\date{}

%----------------------------------------------------------------------------------------

\begin{document}

\maketitle
\newpage

\section{Design Questions: Priority Scheduling}
\subsection{Data Structures}
\subsubsection{Purpose of new variables}

\begin{enumerate}

\item \begin{lstlisting}
struct thread
  {
		.
		.
		.
    struct list locks_holding;           /* List of locks that thread owns. */
    struct lock *waiting_on_lock; 	 /* Lock the thread is waiting on. */ 
    struct semaphore *waiting_on_sema;   /* Semaphore the thread is waiting on. */
		.
		.
		.
  };
\end{lstlisting}

\item \begin{lstlisting}
struct lock 
  {
    struct thread *holder;       /* Thread holding lock (for debugging). */
    struct list_elem lock_elem;  /* For the list of locks a thread has.  */
    struct semaphore semaphore;  /* Binary semaphore controlling access. */
  };
\end{lstlisting}
\end{enumerate}


\begin{enumerate}

\item We have modified \texttt{struct thread} by adding three new members. The first is a list of locks that the thread owns. This 
is useful to modify the effective priority. The second is a pointer to a lock that the thread is waiting on. This pointer is \texttt{NULL}
if the thread is not waiting on any lock. This member is useful to donate priority along the chain. The third is another pointer to a semaphore. This is used in the same manner as the second member but with semaphores, since a thread could be waiting on a semaphore but not on a lock.

\item We have modified only one member in \texttt{struct lock}. This is a \texttt{list\_elem} for the list of locks a thread owns, as explained
in the previous part.
\end{enumerate}



\subsubsection{Data structure used to track priority donation}

\subsection{Algorithms}
\subsubsection{Ensuring highest priority thread wakes up first}

\subsubsection{Sequence of events in \texttt{lock\_acquire()}}

\subsubsection{Sequence of events in \texttt{lock\_release()}}

\subsection{Synchronization}
\subsubsection{Potential race in \texttt{thread\_set\_priority()}}

\subsection{Rationale}

\subsubsection{Choice of design and superiority to other designs}

\newpage

\section{Design Questions: Advanced Scheduler}
\subsection{Data Structures}
\subsubsection{Purpose of new variables}

\subsection{Algorithms}
\subsubsection{Table}
\subsection{Ambiguities in scheduler specification and resolution}
\subsubsection{Dividing cost of scheduling}

\subsection{Rationale}
\subsubsection{Critique of design}

\subsubsection{Design in fixed-point arithmetic}


%----------------------------------------------------------------------------------------

\end{document}